---
format:
  pdf:
    pdf-engine: xelatex
    documentclass: article
    classoption:
      - notitlepage
    mainfont: "Times New Roman"
    toc: false
    toc-depth: 2
    number-sections: true
execute:
  warning: false
  message: false
header-includes:
  - \usepackage{titling}
  - \setlength{\droptitle}{-1.2cm}  
  - \usepackage{float}
  - \usepackage{placeins}
  - \setlength{\textfloatsep}{5pt}
  - \setlength{\floatsep}{5pt}
  - \setlength{\intextsep}{5pt}
  - \usepackage[margin=1in]{geometry}
---

```{=tex}
\begin{center}
{\LARGE Heart Failure Analysis in MIMIC-III\par}
\vspace{0.2em}
{\large `r Sys.Date()`\par}
\end{center}
\vspace{1.5em}
```
This report summarizes several characteristics of heart failure patients in the MIMIC-III dataset.

```{r}
#| echo: false

#load required packages
library(kableExtra)
library(tidyverse)
library(dbplyr)
library(readr)
library(gt)
library(gtsummary)
library(ggplot2)
library(here)

#load data
data  <- read.csv(here("HDMAssessment", "hf_summary_final.csv"))
data2 <- read.csv(here("HDMAssessment", "nonhf_summary_final.csv"))
data3 <- read.csv(here("HDMAssessment", "question2.csv"))
data4 <- read.csv(here("HDMAssessment", "question3.csv"))
```

## Heart Failure vs Non-Heart Failure

Table 1 compares heart failure and non-heart failure hospital admissions by diabetes status, admission type, ethnicity, gender, and age group.

```{r}
#| echo: false
#|
combined_summary <- data %>%
  mutate(hf_status = "Heart Failure") %>%
  bind_rows(
    data2 %>% mutate(hf_status = "Non-Heart Failure")
  )

combined_summary <- combined_summary %>%
  mutate(
    admission_type = fct_recode(admission_type,
                                 "Elective" = "ELECTIVE",
                                 "Urgent" = "URGENT",
                                 "Emergency" = "EMERGENCY",
                                "Newborn" = "NEWBORN"),
    ethnicity_grouped = fct_recode(ethnicity_grouped,
                                   "Asian" = "asian",
                                   "Black" = "black",
                                   "Hispanic" = "hispanic",
                                   "Native" = "native",
                                   "Other" = "other",
                                   "White" = "white"),
    gender = fct_recode(gender,
                        "Female" = "F",
                        "Male" = "M"))

combined_summary <- combined_summary %>%
  mutate(
    ethnicity_grouped = ethnicity_grouped %>%
      fct_explicit_na(na_level = "Unknown") %>%
      fct_recode(
        "Asian" = "asian",
        "Black" = "black",
        "Hispanic" = "hispanic",
        "Native" = "native",
        "Other" = "other",
        "White" = "white",
        "Unknown" = "unknown"
      ) %>%
       fct_relevel(
        "Asian", "Black", "Hispanic", "Native", "Other", "White", "Unknown"
      )
  )

hf_vs_nonhf_table <- combined_summary %>%
  select(
    has_secondary_diabetes,
    admission_type,
    ethnicity_grouped,
    gender,
    age_group,
    hf_status
  ) %>%
  tbl_summary(
    by = hf_status,
     label = list(
      has_secondary_diabetes ~ "Secondary Diabetes",
      admission_type ~ "Admission Type",
      ethnicity_grouped ~ "Ethnicity",
      gender ~ "Gender",
      age_group ~ "Age Group (years)"
    ),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} / {N} ({p}%)"
    ),
    missing = "ifany"
  ) %>%
  bold_labels()

hf_vs_nonhf_table %>%
  as_kable_extra(
    booktabs = TRUE,
    linesep = "",
    caption = "Comparison of demographic and clinical characteristics between heart failure and non–heart failure patients"
  ) %>%
  kable_styling(
    latex_options = c("scale_down", "HOLD_position"),  
    font_size = 9
  )
```

\small Note: Patients with secondary diabetes both with and without complications were categorized as having diabetes.

Key takeaways: (1) Admissions are more likely to be an emergency if a patient has heart failure compared to one who does not. (2) Heart failure admissions are more common in the later age groups compared to non-heart failure admissions.

\vspace{5em}

## Heart Failures - Time in Hospital, ICU, & Death Counts

Tables 2-5 summarize information about heart failure patients' deaths and overall time in the hospital & ICU, each stratified by a different characteristic. Patients without known information about the stratifications were removed from the corresponding table.

```{r}
#| echo: false
#summary stratified by age
age_table <- data3 %>%
  select(
    age_strat,
    time_in_hospital,
    time_in_icu,
    died
  ) %>%
  tbl_summary(
    by = age_strat,
    label = list(
      time_in_hospital ~ "Time in hospital (days)",
      time_in_icu ~ "Time in ICU (days)",
      died ~ "Died"
    ),
    statistic = list(
      time_in_hospital ~ "{median} ({p25}, {p75})",
      time_in_icu ~ "{median} ({p25}, {p75})",
      died ~ "{n} / {N} ({p}%)"
    ),
    missing = "ifany"
  ) %>%
  modify_header(label ~ "**Age group (years)**") %>%
  modify_footnote(
    all_continuous() ~ "Median (25th percentile, 75th percentile)"
  ) %>%
  bold_labels()

age_table %>%
  as_kable_extra(booktabs = TRUE, linesep = "",
                 caption = "Summary of hospital and ICU length of stay and in-hospital mortality, stratified by age") %>%
  kable_styling(latex_options = c("scale_down",  "HOLD_position"),
                font_size = 9)
```

\small Note: No heart failure admissions were age 0-17, so this stratum was omitted. \vspace{2em}

```{r}
#| echo: false
#summary stratified by gender
gender_table <- data3 %>%
  mutate(gender = recode(gender,
                         "F" = "Female",
                         "M" = "Male")) %>%
  select(
    gender,
    time_in_hospital,
    time_in_icu,
    died
  ) %>%
  tbl_summary(
    by = gender,
    label = list(
      time_in_hospital ~ "Time in hospital (days)",
      time_in_icu ~ "Time in ICU (days)",
      died ~ "Died"
    ),
    statistic = list(
      time_in_hospital ~ "{median} ({p25}, {p75})",
      time_in_icu ~ "{median} ({p25}, {p75})",
      died ~ "{n} / {N} ({p}%)"
    ),
    missing = "ifany"
  ) %>%
  modify_header(label ~ "**Gender**") %>%
  add_overall() %>%
  bold_labels()

gender_table %>%
  as_kable_extra(booktabs = TRUE, linesep = "",
                 caption = "Summary of hospital and ICU length of stay and in-hospital mortality, stratified by gender") %>%
  kable_styling(latex_options = c("scale_down", "HOLD_position"),
                font_size = 9)
```

\vspace{2em}

```{r}
#| echo: false


library(stringr)

ethnicity_table <- data3 %>%
  mutate(ethnicity_grouped = str_to_title(ethnicity_grouped)) %>%  # Capitalize each word
  select(
    ethnicity_grouped,
    time_in_hospital,
    time_in_icu,
    died
  ) %>%
  tbl_summary(
    by = ethnicity_grouped,
    label = list(
      time_in_hospital ~ "Time in hospital (days)",
      time_in_icu ~ "Time in ICU (days)",
      died ~ "Died"
    ),
    statistic = list(
      time_in_hospital ~ "{median} ({p25}, {p75})",
      time_in_icu ~ "{median} ({p25}, {p75})",
      died ~ "{n} / {N} ({p}%)"
    ),
    missing = "ifany"
  ) %>%
  bold_labels() %>%
  modify_header(label ~ "**Ethnicity**") %>%  # Change "Characteristic" to "Ethnicity"
  modify_footnote(
    all_continuous() ~ "Median (25th percentile, 75th percentile)"
  )

ethnicity_table %>%
  as_kable_extra(booktabs = TRUE, linesep = "",
                 caption = "Summary of hospital and ICU length of stay and in-hospital mortality, stratified by ethnicity") %>%
  kableExtra::kable_styling(latex_options = c("scale_down", "HOLD_position"), font_size = 11)

```

\vspace{2em}

```{r}
#| echo: false
#summary stratified by diabetes status
diabetes_table <- data3 %>%
  select(
    has_secondary_diabetes,
    time_in_hospital,
    time_in_icu,
    died
  ) %>%
  tbl_summary(
    by = has_secondary_diabetes,
      label = list(
      time_in_hospital ~ "Time in hospital (days)",
      time_in_icu ~ "Time in ICU (days)",
      died ~ "Died"),
    statistic = list(
      time_in_hospital ~ "{median} ({p25}, {p75})",
      time_in_icu ~ "{median} ({p25}, {p75})",
      died ~ "{n} / {N} ({p}%)"
    ),
    missing = "ifany"
  ) %>%
  bold_labels() %>%
  modify_header(label ~ "**Has Secondary Diabetes**")

diabetes_table %>%
  as_kable_extra(booktabs = TRUE, linesep = "",
                 caption = "Summary of hospital and ICU length of stay and in-hospital mortality, stratified by diabetes status") %>%
  kable_styling(latex_options = c("scale_down", "HOLD_position"),
                font_size = 9)

```

Key takeaway: Days spent in the hospital/ICU and death counts are not vastly different between any of the above stratifications.

\vspace{3em}

## Heart Failures - Prescription Characteristics

Table 6 and Figure 1 both refer to information regarding top prescriptions among heart failure patients.

```{r}
#| echo: false
#|results: asis

library(dplyr)
library(tidyr)
library(kableExtra)
library(ggplot2)

# Summarize drug data
drug_table <- data4 %>%
  group_by(drug) %>%
  summarise(
    total_prescriptions = n(),
    avg_dose = mean(as.numeric(dose_val_rx), na.rm = TRUE),
    icu_prescriptions = sum(!is.na(icustay_id)),
    avg_prescriptions_per_admission = n() / n_distinct(hadm_id),
    dose_unit = first(dose_unit_rx),
    .groups = "drop"
  ) %>%
  mutate(avg_dose_with_unit = paste0(round(avg_dose, 2), " ", dose_unit),
        avg_prescriptions_per_admission = round(avg_prescriptions_per_admission, 2))

```

\FloatBarrier

::: columns
```{r tbl6, echo=FALSE, results="asis"}
drug_table %>%
  dplyr::select(drug, avg_dose_with_unit, avg_prescriptions_per_admission) %>%
  kableExtra::kbl(
    booktabs = TRUE,
    col.names = c("Drug", "Average Dose", "Average Number of Prescriptions per Admission"),
    caption = "Average doses of top five prescriptions in heart failure patients"
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("HOLD_position"),
    font_size = 10
  )
```
:::

```{r, echo=FALSE}
#| fig-cap: "Total vs ICU prescriptions for the top five drugs"
#| fig-pos: "H"
#| out-width: "100%"

#pivot table so that total and icu prescriptions can be compared
drug_table_long <- drug_table %>%
dplyr::select(drug, total_prescriptions, icu_prescriptions) %>%
tidyr::pivot_longer(
cols = c(total_prescriptions, icu_prescriptions),
names_to = "type",
values_to = "count"
) %>% mutate(
  type = recode(type,
                total_prescriptions = "All prescriptions",
                icu_prescriptions = "ICU prescriptions"))

ggplot(drug_table_long, aes(x = drug, y = count, fill = type)) +
geom_col(position = "dodge") +
labs(x = "Drug", y = "Number of Prescriptions", fill = "Prescription Type") +
theme_classic() +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Key takeaways: (1) The most frequently prescribed medication both in the hospital as a whole and within the ICU is Furosemide. (2) The average number of prescriptions per admission does not largely differ between each of the five drugs.

\FloatBarrier